---
title: "FinancialAnalysis"
author: "Toan Truong Ho Nguyen"
format: html
---

# Financial Analysis Final Project

## The Key Question:

#### How are investor age, gender, expected return, and investment duration associated with preference for equity-market investments?

## Set Up

Load the data set

```{r}
finance_trends <- read.csv("Finance_Trends.csv")
finance_trends
```

### Question 1:

Which investment avenues receive the strongest average rankings, and how much variation exists in each investment's ranking?

```{r}
library(tidyverse)
finance_trends <- read_csv("Finance_Trends.csv")
dim(finance_trends)
```

```{r}
finance_trends |> select(
  Mutual_Funds,
  Equity_Market,
  Debentures,
  Government_Bonds,
  Fixed_Deposits,
  PPF,
  Gold
) |>
  summary()
```

Ranking varibles:

```{r}
investment_rankings <- finance_trends |> select(
   Mutual_Funds,
   Equity_Market,
   Debentures,
   Government_Bonds,
   Fixed_Deposits,
   PPF,
   Gold
)
glimpse(investment_rankings)
```

```{r}
investment_rankings_long <- investment_rankings |> pivot_longer(
  cols = everything(),
  names_to = "investment",
  values_to = "rank") |> mutate(investment = str_replace_all(investment, "_", " "))

stopifnot(
  all(investment_rankings_long$rank >= 1),
  all(investment_rankings_long$rank <= 7),
  !any(is.na(investment_rankings_long$rank))
)

ranking_summary <- investment_rankings_long |>
  group_by(investment) |> summarise(
    n = n(),
    mean_rank = mean(rank),
    median_rank = median(rank),
    sd_rank = sd(rank),
    .groups = "drop"
  ) |> arrange(mean_rank)

ranking_summary_display <- ranking_summary |> mutate(
   mean_rank = round(mean_rank, 2),
   median_rank = round(median_rank, 2),
   sd_rank = round(sd_rank, 2))

ranking_summary_display
```

```{r}
ggplot(
  data = ranking_summary,
  mapping = aes(
    x = reorder(investment, mean_rank),
    y = mean_rank)
) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(
    breaks = 1:7,
    limits = c(0, 7)) +
  labs(
    title = "Average Ranking of Investment Avenues",
    subtitle = "Lower average rankings indicate stronger preferences",
    x = "Investment Avenue",
    y = "Average Ranking") +
  theme_minimal()
```

### Question 2:
Does the percentage of respondents who rank Equity Market among their top two investment choices differ between men and women?

```{r}
equity_gender <- finance_trends |>
  select(
    gender,
    Equity_Market
  ) |> mutate(
    equity_top2 = if_else(
      Equity_Market <= 2,
      "Yes",
      "No"),
     gender = factor(gender),
    equity_top2 = factor(equity_top2))

table(
  equity_gender$gender,
  equity_gender$equity_top2)

gender_summary <- equity_gender |>
  group_by(gender, equity_top2) |>
  summarise(
    n = n(),
    .groups = "drop") |> 
  group_by(gender) |> mutate(
     percent = n / sum(n) * 100 )

gender_summary
    
```

```{r}
ggplot(
  gender_summary,
  aes(
    x = gender,
    y = percent,
    fill = equity_top2)
) +
  geom_col(position = "dodge") +
  labs(
    title = "Equity Market Top-Two Preference by Gender",
    x = "Gender",
    y = "Percent of Respondents",
    fill = "Top Two Ranking") + theme_minimal()

gender_equity_test <- chisq.test(
  table(
    equity_gender$gender,
     equity_gender$equity_top2)
)

gender_equity_test
```


























